Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-07 WSSSPE3/report/WSSSPE3_review.tex
fix a lot of typos, mostly added by myself

Turk talked about possible reasons why we want to sustain scientific
software: devotion to science and interests in pursuing the next stage of
research; fun and creative thrill in writing codes and papers; usefulness with
measurable impacts, for example, LINPACK and HDF groups providing of data
storage to satellites, which goes beyond usefulness to necessity. Lastly, Turk
presented his wishlist of questions to be answered in the future: 
%
\begin{itemize}

@@

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces to
Improve Productivity of Legacy Tools}}
Abani Patra presented an example how the community using Titan2D, a Geoflow
Simulation Software, increased the productivity of their tools by improving
both code and data layout. The main obstacles in this change were the non-existence
of a common version control system for the source code, coupled with multiple
versions of the same code base, the fixed format of input files, that many
input values used to be set as compilation flags, and that the internal data
layout was not suitable for modern technologies (vectorization, accelerators).
The approach of the Titan2D developers included reinforcing the code structure
using a multiple laters of Python and C++, and a redesign of the data layout to
be better suitable for modern CPUs and accelerators.

@@

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham and
Sergei Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
Biology: Lessons Learned on the KBase Project}}
In this talk, Dan Gunter presented experiences and lessons learned as part of
KBase, an open software and data platform for addressing the grand challenge of
systems biology: predicting and designing biological function. KBase is a
unified system that integrates data and analytical tools for comparative
functional genomics of microbes, plants, and their communities. However,
it is also a collaborative environment for sharing methods and results, and
placing those results in the context of knowledge in the field. Being a large,
multiinstutional project, one the big challenges is to agree on standard to
enable a single, maintainable system. Working in isolation does not work (anymore)
within this field, and no standards for software engineering existed either.
This is contrasted to computer science research, where these approaches can lead to
shorter design cycles, leading to more time for highly rewarded activities like
publishing, performance studies, graduating students, or protecting ideas
before publication. Sustainability is not of big concern there, compared to
these. Instead of this more traditional approach, the following is used within
KBase for large, distributed teams. After picking projects and team members, a
number of 4-5 teams uses about two weeks to work on their project before a
1-week evaluation by the executive committee, and after which new teams and
projects might be chosen to restart the ``agile'' development cycle. This
process is intentionally open and documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}}
Yolanda Gil presented an approach to overcome some hurdles in scientific
publishing: the disconnect between experimental data, research software, and
publications. A current effort in the Geosciences is the ``Geoscience Paper of
the Future'', which includes four elements. Firstly, it forms a modern paper
including both text, data and pointers to supplementary matertial. Secondly, is
reproducible, including data processing, workflow and visualization tools. As
third element, it is part of open science, which includes being publicly
available under open licenses, and providing meta-data. Last, but not least, it
uses digital scholarship elements like persisent identifiers for software,
data, and authors, and it cites both data and software. OntoSoft is an effort
to help this effort, providing software stewardship for the Geosciences. As
part of this inituative, a special issue of a journal in a geoscience areas is
planned that only includes geoscience papers of the future, with submissions
open until the end of 2015. In addition, training sessions are provided to
geoscientists to learn best practices in software and data sharing, provenance
documentation, and scholarly publication.

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software Accreditation
Framework}}
Neil Chue Hong presented a proposal to build a scientific software
accreditation framework. One of the aims of such a framework would be to
measure how ``good' a given software is, and to evaluate how this can be
effectively measured in the first place. This can be compared to the effort of
the standardized, easy to read and understand nutritional labelling of goods,
which only contain a small set of categories. However, such a framework for
software would have to be able to evaluate differently due to different
existing community norms. The challenges such a framework faces include that
many measurements are subjective, many metrics are too costly, and that
self-assessment needs to be encouraged. Possible categories would include
availability, usability, transferability, among others. Such a framework could
enable both improvement of specific software, as well as comparisons of similar
software. An accreditation by such a framework could then be part of software
management plans, ensuring that software is accessible and reusable throughout
the proposed project and beyond.

@@

\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
included)}}
Alice Allen described the Astrophysics Source Code Library (ascl.net), an increasingly
used way to obtain a unique ID for astrophysics software that is indexed by indexing
services and can be cited. ASCL offers clones of existing infrastructure, provides server
space and computing resources, shares innovations, and permits moves elsewhere. In turn,
users have to provide a domain nam, control and configure their site and use it as
intended, gather their codes to their liking, share innovations and protect the
provided computing environment.

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While Developing
Community}}

@@

The workshop had two components, presentations and working groups.  The presentations,
in the first half day of the workshop, including a inspirational keynote and a set of lightning
talks.
We used lightning talks for two reasons: first, the need of some participants to have a slot
on the agenda to justify their attendance, and second, as a way to get new ideas across
to all the attendees.  We broke with the tradition of requiring the lightning talk submitters
to self-publish their papers, and instead used a common peer-review platform\footnote{\url{http://easychair.org}}, choosing to publish the speaker slides on the workshop web site
instead.

>>>>

Turk talked about possible reasons why we want to sustain scientific
software: devotion to science and interests in pursuing the next stage of
research; fun and creative thrill in writing codes and papers; usefulness with
measurable impacts, for example, LINPACK and HDF groups providing of data
storage to satellites, which goes beyond usefulness to necessity. Lastly, Turk
presented his wish-list of questions to be answered in the future: 
%
\begin{itemize}

@@

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces to
Improve Productivity of Legacy Tools}}
Abani Patra presented an example how the community using Titan2D, a Geoflow
Simulation Software, increased the productivity of their tools by improving
both code and data layout. The main obstacles in this change were the non-existence
of a common version control system for the source code, coupled with multiple
versions of the same code base, the fixed format of input files, that many
input values used to be set as compilation flags, and that the internal data
layout was not suitable for modern technologies (vectorization, accelerators).
The approach of the Titan2D developers included reinforcing the code structure
using a multiple layers of Python and C++, and a redesign of the data layout to
be better suitable for modern CPUs and accelerators.

@@

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham and
Sergei Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
Biology: Lessons Learned on the KBase Project}}
In this talk, Dan Gunter presented experiences and lessons learned as part of
KBase, an open software and data platform for addressing the grand challenge of
systems biology: predicting and designing biological function. KBase is a
unified system that integrates data and analytical tools for comparative
functional genomics of microbes, plants, and their communities. However,
it is also a collaborative environment for sharing methods and results, and
placing those results in the context of knowledge in the field. Being a large,
multi-institutional project, one the big challenges is to agree on standard to
enable a single, maintainable system. Working in isolation does not work (anymore)
within this field, and no standards for software engineering existed either.
This is contrasted to computer science research, where these approaches can lead to
shorter design cycles, leading to more time for highly rewarded activities like
publishing, performance studies, graduating students, or protecting ideas
before publication. Sustainability is not of big concern there, compared to
these. Instead of this more traditional approach, the following is used within
KBase for large, distributed teams. After picking projects and team members, a
number of 4-5 teams uses about two weeks to work on their project before a
1-week evaluation by the executive committee, and after which new teams and
projects might be chosen to restart the ``agile'' development cycle. This
process is intentionally open and documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}}
Yolanda Gil presented an approach to overcome some hurdles in scientific
publishing: the disconnect between experimental data, research software, and
publications. A current effort in the Geosciences is the ``Geoscience Paper of
the Future'', which includes four elements. Firstly, it forms a modern paper
including both text, data and pointers to supplementary material. Secondly, is
reproducible, including data processing, workflow and visualization tools. As
third element, it is part of open science, which includes being publicly
available under open licenses, and providing meta-data. Last, but not least, it
uses digital scholarship elements like persistent identifiers for software,
data, and authors, and it cites both data and software. OntoSoft is an effort
to help this effort, providing software stewardship for the Geosciences. As
part of this initiative, a special issue of a journal in a Geoscience areas is
planned that only includes Geoscience papers of the future, with submissions
open until the end of 2015. In addition, training sessions are provided to
geoscientists to learn best practices in software and data sharing, provenance
documentation, and scholarly publication.

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software Accreditation
Framework}}
Neil Chue Hong presented a proposal to build a scientific software
accreditation framework. One of the aims of such a framework would be to
measure how ``good' a given software is, and to evaluate how this can be
effectively measured in the first place. This can be compared to the effort of
the standardized, easy to read and understand nutritional labeling of goods,
which only contain a small set of categories. However, such a framework for
software would have to be able to evaluate differently due to different
existing community norms. The challenges such a framework faces include that
many measurements are subjective, many metrics are too costly, and that
self-assessment needs to be encouraged. Possible categories would include
availability, usability, transferability, among others. Such a framework could
enable both improvement of specific software, as well as comparisons of similar
software. An accreditation by such a framework could then be part of software
management plans, ensuring that software is accessible and reusable throughout
the proposed project and beyond.

@@

\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
included)}}
Alice Allen described the Astrophysics Source Code Library (ascl.net), an
increasingly used way to obtain a unique ID for astrophysics software that is
indexed by indexing services and can be cited. ASCL offers clones of existing
infrastructure, provides server space and computing resources, shares
innovations, and permits moves elsewhere. In turn, users have to provide a
domain name, control and configure their site and use it as intended, gather
their codes to their liking, share innovations and protect the provided
computing environment.

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While Developing
Community}}
The effort behind this talk started out as break-out team of an NSF SI2
workshop in 2015, and centers around creating a framework for creation of
metrics for scientific software. This framework would improve both the metrics,
and the software it evaluates, and tries to build a community around this idea.
With especially the latter idea in mind, a software ``peer review group'' would
be created, representing stakeholders who will self-review software created by
their respective communities, and will concurrently develop metrics. The whole
project should be community-governed, without a single institution overseeing
the activity or infrastructure, with the hope to evolving community-generated
and adopted standards. The generation of metrics would be tied to the actual
evaluation of software, creating incentive by improving the evaluated software
itself during this process. Infrastructure for the creation of metrics and
evaluation would be provided by the framework code itself, but also provided
code reviews of the evaluated software, as well as socially through forums for
generation of software success metrics. With enough examples one of the open
questions is whether it is possible to fit the resulting metrics in a common
template. So far, this is still in a design phase, with a white paper at the
CSESSP workshop 2015 and this talk, but WSSSPE workshop are seen as a forum for
the community to assemble and act, and are planned to be used also in the
future to build this community and framework.

@@

The workshop had two components, presentations and working groups.  The presentations,
in the first half day of the workshop, including a inspirational keynote and a set of lightning
talks.
We used lightning talks for two reasons: first, the need of some participants to have a slot
on the agenda to justify their attendance, and second, as a way to get new ideas across
to all the attendees.  We broke with the tradition of requiring the lightning talk submitter
to self-publish their papers, and instead used a common peer-review platform\footnote{\url{http://easychair.org}}, choosing to publish the speaker slides on the workshop web site
instead.
